---
title: "Introduction to graphlayouts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to graphlayouts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6,
  fig.align = 'center'
)
```

The package implements some graph layout algorithms that are not available
in the `igraph` package. 

The package implements three algorithms which serve different purposes:

- Stress majorization for general layouts ([Paper](https://graphviz.gitlab.io/_pages/Documentation/GKN04.pdf))
- Quadrilateral backbone layout for (potentially) clustered networks ([Paper](http://jgaa.info/accepted/2015/NocajOrtmannBrandes2015.19.2.pdf))
- flexible radial layouts to focus on specific nodes (e.g. ego in networks) ([Paper](http://jgaa.info/accepted/2011/BrandesPich2011.15.1.pdf))
- spectral layouts based on eigenvectors of matrices

## Stress Majorization: Connected Network  

*This example is a bit of a special case since it exploits some weird issues in igraph.* 
```{r example}
library(igraph)   
library(ggraph)   
library(graphlayouts)

set.seed(666)
pa <- sample_pa(1000,1,1,directed = F)

ggraph(pa)+
  geom_edge_link(width=0.2,colour="grey")+
  geom_node_point(col="black",size=0.3)+
  theme_graph()

ggraph(pa,layout="stress")+
  geom_edge_link(width=0.2,colour="grey")+
  geom_node_point(col="black",size=0.3)+
  theme_graph()
```

## Stress Majorization: Unconnected Network

Stress majorization also works for networks with several components. It relies 
on a bin packing algorithm to efficiently put the components in a rectangle, rather than a circle.

```{r example_un}
set.seed(666)
g <- disjoint_union(
  sample_pa(10,directed = F),
  sample_pa(20,directed = F),
  sample_pa(30,directed = F),
  sample_pa(40,directed = F),
  sample_pa(50,directed = F),
  sample_pa(60,directed = F),
  sample_pa(80,directed = F)
)

ggraph(g) +
  geom_edge_link() +
  geom_node_point() +
  theme_graph()

ggraph(g, layout="stress") +
  geom_edge_link() +
  geom_node_point() +
  theme_graph()
```

## Backbone Layout

Backbone layouts are helpful for drawing so called hairballs. 

```{r hairball}
set.seed(665)
#create network with a group structure
g <- sample_islands(9,40,0.4,15)
g <- simplify(g)
V(g)$grp <- as.character(rep(1:9,each=40))

ggraph(g,layout="stress")+
  geom_edge_link(colour=rgb(0,0,0,0.5),width=0.1)+
  geom_node_point(aes(col=grp))+
  scale_color_brewer(palette = "Set1")+
  theme_graph()+
  theme(legend.position = "none")

```


The backbone layout helps to uncover potential group structures based on edge 
embeddedness and puts more emphasis on this structure in the layout.

```{r backbone}
bb <- layout_as_backbone(g,keep=0.4)
E(g)$col <- F
E(g)$col[bb$backbone] <- T

ggraph(g,layout="manual",node.positions=data.frame(x=bb$xy[,1],y=bb$xy[,2]))+
  geom_edge_link(aes(col=col),width=0.1)+
  geom_node_point(aes(col=grp))+
  scale_color_brewer(palette = "Set1")+
  scale_edge_color_manual(values=c(rgb(0,0,0,0.3),rgb(0,0,0,1)))+
  theme_graph()+
  theme(legend.position = "none")

```


## Radial Layout with Focal Node

The function `layout_with_focus` creates a radial layout around a focal node.
All nodes with the same distance from the focal node are on the same circle.

```{r karate, include=FALSE}
el <- structure(c("Mr Hi", "Mr Hi", "Mr Hi", "Mr Hi", "Mr Hi", "Mr Hi", 
"Mr Hi", "Mr Hi", "Mr Hi", "Mr Hi", "Mr Hi", "Mr Hi", "Mr Hi", 
"Mr Hi", "Mr Hi", "Mr Hi", "Actor 2", "Actor 2", "Actor 2", "Actor 2", 
"Actor 2", "Actor 2", "Actor 2", "Actor 2", "Actor 3", "Actor 3", 
"Actor 3", "Actor 3", "Actor 3", "Actor 3", "Actor 3", "Actor 3", 
"Actor 4", "Actor 4", "Actor 4", "Actor 5", "Actor 5", "Actor 6", 
"Actor 6", "Actor 6", "Actor 7", "Actor 9", "Actor 9", "Actor 9", 
"Actor 10", "Actor 14", "Actor 15", "Actor 15", "Actor 16", "Actor 16", 
"Actor 19", "Actor 19", "Actor 20", "Actor 21", "Actor 21", "Actor 23", 
"Actor 23", "Actor 24", "Actor 24", "Actor 24", "Actor 24", "Actor 24", 
"Actor 25", "Actor 25", "Actor 25", "Actor 26", "Actor 27", "Actor 27", 
"Actor 28", "Actor 29", "Actor 29", "Actor 30", "Actor 30", "Actor 31", 
"Actor 31", "Actor 32", "Actor 32", "Actor 33", "Actor 2", "Actor 3", 
"Actor 4", "Actor 5", "Actor 6", "Actor 7", "Actor 8", "Actor 9", 
"Actor 11", "Actor 12", "Actor 13", "Actor 14", "Actor 18", "Actor 20", 
"Actor 22", "Actor 32", "Actor 3", "Actor 4", "Actor 8", "Actor 14", 
"Actor 18", "Actor 20", "Actor 22", "Actor 31", "Actor 4", "Actor 8", 
"Actor 9", "Actor 10", "Actor 14", "Actor 28", "Actor 29", "Actor 33", 
"Actor 8", "Actor 13", "Actor 14", "Actor 7", "Actor 11", "Actor 7", 
"Actor 11", "Actor 17", "Actor 17", "Actor 31", "Actor 33", "John A", 
"John A", "John A", "Actor 33", "John A", "Actor 33", "John A", 
"Actor 33", "John A", "John A", "Actor 33", "John A", "Actor 33", 
"John A", "Actor 26", "Actor 28", "Actor 30", "Actor 33", "John A", 
"Actor 26", "Actor 28", "Actor 32", "Actor 32", "Actor 30", "John A", 
"John A", "Actor 32", "John A", "Actor 33", "John A", "Actor 33", 
"John A", "Actor 33", "John A", "John A"), .Dim = c(78L, 2L))

karate <- graph_from_edgelist(el,directed = FALSE)
V(karate)$Faction <- c(1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 1, 1, 1, 1, 2, 2, 1, 1, 2, 1, 
2, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2)
```


```{r flex_focus}
#library(igraphdata)
#data("karate")

p1 <- ggraph(karate,layout = "focus",v = 1) +
  draw_circle(use = "focus",max.circle = 3)+
  geom_edge_link(edge_color="black",edge_width=0.3)+
  geom_node_point(aes(fill=as.factor(Faction)),size=2,shape=21)+
  scale_fill_manual(values=c("#8B2323", "#EEAD0E"))+
  theme_graph()+
  theme(legend.position = "none")+
  coord_fixed()+
  labs(title= "Focus on Mr. Hi")

p2 <- ggraph(karate,layout = "focus",v = 34) +
  draw_circle(use = "focus",max.circle = 4)+
  geom_edge_link(edge_color="black",edge_width=0.3)+
  geom_node_point(aes(fill=as.factor(Faction)),size=2,shape=21)+
  scale_fill_manual(values=c("#8B2323", "#EEAD0E"))+
  theme_graph()+
  theme(legend.position = "none")+
  coord_fixed()+
  labs(title= "Focus on John A.")

p1
p2
```


## Radial Centrality Layout

The function `layout_with_centrality` creates a radial layout around the node with the
highest centrality value. The further outside a node is, the more peripheral it is.

```{r flex_cent}

bc <- betweenness(karate)
p1 <- ggraph(karate,layout = "centrality", cent = bc, tseq = seq(0,1,0.15)) +
  draw_circle(use = "cent") +
  annotate_circle(bc,format="",pos="bottom") +
  geom_edge_link(edge_color="black",edge_width=0.3)+
  geom_node_point(aes(fill=as.factor(Faction)),size=2,shape=21)+
  scale_fill_manual(values=c("#8B2323", "#EEAD0E"))+
  theme_graph()+
  theme(legend.position = "none")+
  coord_fixed()+
  labs(title="betweenness centrality")


cc <- closeness(karate)
p2 <- ggraph(karate,layout = "centrality", cent = cc, tseq = seq(0,1,0.2)) +
  draw_circle(use = "cent") +
  annotate_circle(cc,format="scientific",pos="bottom") +
  geom_edge_link(edge_color="black",edge_width=0.3)+
  geom_node_point(aes(fill=as.factor(Faction)),size=2,shape=21)+
  scale_fill_manual(values=c("#8B2323", "#EEAD0E"))+
  theme_graph()+
  theme(legend.position = "none")+
  coord_fixed()+
  labs(title="closeness centrality")

p1
p2
```

## Spectral Layouts

Spectral layouts use eigenvectors of matrices associated with a graph for layouting.
Possible matrices are the adjacency matrix and the laplacian matrix. Additionally, 
one can choose if the largest or smallest eigenvectors should be used.
```{r eigen_layout}
g <- sample_gnp(50,0.2)

ggraph(g,layout="eigen",type="adjacency",ev="largest")+
  geom_edge_link(n=2,edge_colour="grey66")+
  geom_node_point(shape=21,fill="grey25",size=5)+
  theme_graph()

ggraph(g,layout="eigen",type="adjacency",ev="smallest")+
  geom_edge_link(n=2,edge_colour="grey66")+
  geom_node_point(shape=21,fill="grey25",size=5)+
  theme_graph()


ggraph(g,layout="eigen",type="laplacian",ev="largest")+
  geom_edge_link(n=2,edge_colour="grey66")+
  geom_node_point(shape=21,fill="grey25",size=5)+
  theme_graph()

ggraph(g,layout="eigen",type="laplacian",ev="smallest")+
  geom_edge_link(n=2,edge_colour="grey66")+
  geom_node_point(shape=21,fill="grey25",size=5)+
  theme_graph()
```

